---
title: "tidyHeatmap"
output: github_document
---

(If you like tidyverse and RNA, try [tidybulk](https://github.com/stemangiola/tidybulk) for tidy and modular transcriptomics analyses!)

<!-- badges: start -->
  [![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->


```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, cache.lazy = FALSE)

library(magrittr)
library(dplyr)
library(tidyr)
library(tidyHeatmap)
library(purrr)

```

Tidy heatmap. This package is a tidy wrapper of the package [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html). The goal of this package is to interface tidy data frames with this powerful tool.

Some of the advantages are:

- Row and/or columns colour annotations are easy to integrate just specifying one parameter (column names). 
- Custom grouping of rows is easy to specify providing a grouped tbl. For example `df %>% group_by(...)`
- Labels size adjusted by row and column total number
- Default use of Brewer and Viridis palettes



## Installation

To install the most up-to-date version

```{r, eval=FALSE}

devtools::install_github("stemangiola/tidyHeatmap")

```

To install the most stable version (however please keep in mind that this package is under a maturing lifecycle stage)

```{r, eval=FALSE}

install.packages("tidyHeatmap")

```

## Input data frame

```{r}
mtcars_tidy = 
	mtcars %>% 
	as_tibble(rownames="Car name") %>% 
	
	# Scale
	mutate_at(vars(-`Car name`, -hp, -vs), scale) %>%
	
	# tidyfy
	gather(Property, Value, -`Car name`, -hp, -vs)

mtcars_tidy
```


## Plot

For plotting, you simply pipe the input data frame into heatmap, specifying:

- The rows, cols relative column names (mandatory)
- The value column name (mandatory)
- The annotations column name(s)

mtcars
```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		annotation = hp
	)
```

## Grouping

We can easily group the data (one group per dimension maximum, at the moment only the vertical dimension is supported) with dplyr, and the heatmap will be grouped accordingly

```{r}
mtcars_tidy %>% 
	group_by(vs) %>%
	heatmap(
		`Car name`, 
		Property, 
		Value,
		annotation = hp
	)
```

## Custom palettes

We can easily use custom palette, using strings, hexadecimal color character vector, 

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = c("red", "white", "blue")
	)
```

Or a grid::colorRamp2 functionfor higher flexibility

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = circlize::colorRamp2(c(-2, -1, 0, 1, 2), viridis::magma(5))
	)
```

## Multiple groupings and annotations

```{r}
tidyHeatmap::pasilla %>%
	group_by(location, type) %>%
	heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`,
			annotation = c(condition, activation)
		)
```

## Annotation types
"tile" (default), "point", "bar" and "line" are available

```{r}
# Chreate some more data points
pasilla_plus = 
	tidyHeatmap::pasilla %>%
		dplyr::mutate(act = activation) %>% 
		tidyr::nest(data = -sample) %>%
		dplyr::mutate(size = rnorm(n(), 4,0.5)) %>%
		dplyr::mutate(age = runif(n(), 50, 200)) %>%
		tidyr::unnest(data) 

# Plot
pasilla_plus %>%
		tidyHeatmap::heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`,
			annotation = c(condition, activation, act, size, age),
			type = c("tile", "point", "tile", "bar", "line")
		)
```