---
title: "tidyHeatmap"
output: github_document
---

```{r echo=FALSE}
knitr::opts_chunk$set( fig.path = "man/figures/")
```
<!-- badges: start -->
[![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html)
[![DOI](https://joss.theoj.org/papers/10.21105/joss.02472/status.svg)](https://doi.org/10.21105/joss.02472)
<!-- badges: end -->

## Citation
Mangiola et al., (2020). tidyHeatmap: an R package for modular heatmap production based on tidy principles. Journal of Open Source Software, 5(52), 2472, https://doi.org/10.21105/joss.02472

Please have a look also to 

- [nanny](https://github.com/stemangiola/nanny/) for tidy high-level data analysis and manipulation 
- [tidygate](https://github.com/stemangiola/tidygate/) for adding custom gate information to your tibble 
- [tidySingleCellExperiment](https://stemangiola.github.io/tidySingleCellExperiment/) for tidy manipulation of Seurat objects
- [tidyseurat](https://stemangiola.github.io/tidyseurat/) for tidy manipulation of Seurat objects
- [tidybulk](https://stemangiola.github.io/tidybulk/) for tidy high-level data analysis and manipulation 
- [tidySummarizedExperiment](https://stemangiola.github.io/tidySummarizedExperiment/) for heatmaps produced with tidy principles

website: [stemangiola.github.io/tidyHeatmap](https://stemangiola.github.io/tidyHeatmap/)

`tidyHeatmap` is a package that introduces tidy principles to the creation of information-rich heatmaps. 
This package uses [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html) as graphical engine.

**Advantages:**

  - Modular annotation with just specifying column names
  - Custom grouping of rows is easy to specify providing a grouped tbl.
    For example `df %>% group_by(...)`
  - Labels size adjusted by row and column total number
  - Default use of Brewer and Viridis palettes

## Functions/utilities available

Function | Description
------------ | -------------
`heatmap` | Plot base heatmap
`add_tile` | Add tile annotation to the heatmap
`add_point` | Add point annotation to the heatmap
`add_bar` | Add bar annotation to the heatmap
`add_line` | Add line annotation to the heatmap
`layer_point` | Add layer of symbols on top of the heatmap
`layer_square` | Add layer of symbols on top of the heatmap
`layer_diamond` | Add layer of symbols on top of the heatmap
`layer_arrow_up` | Add layer of symbols on top of the heatmap
`layer_arrow_down` | Add layer of symbols on top of the heatmap
`save_pdf` | Save the PDF of the heatmap

## Installation

To install the most up-to-date version

```{r, eval=FALSE}

devtools::install_github("stemangiola/tidyHeatmap")


```


To install the most stable version (however please keep in mind that this package is under a maturing lifecycle stage)

```{r, eval=FALSE}

install.packages("tidyHeatmap")

```

## Contribution

If you want to contribute to the software, report issues or problems with the software or seek support please open an issue [here](https://github.com/stemangiola/tidyHeatmap/issues)

## Input data frame

The heatmaps visualise a multi-element, multi-feature dataset, annotated with independent variables. Each observation is a element-feature pair (e.g., person-physical characteristics).

element | feature | value | independent_variables
------------ | ------------- | ------------- | -------------
`chr` or `fctr` | `chr` or `fctr` | `numeric` | ...

Let's transform the mtcars dataset into a tidy "element-feature-independent variables" data frame. Where the independent variables in this case are 'hp' and 'vs'.

```{r, echo=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(tidyHeatmap)
```

```{r}
mtcars_tidy <- 
	mtcars %>% 
	as_tibble(rownames="Car name") %>% 
	
	# Scale
	mutate_at(vars(-`Car name`, -hp, -vs), scale) %>%
	
	# tidyfy
	pivot_longer(cols = -c(`Car name`, hp, vs), names_to = "Property", values_to = "Value")

mtcars_tidy
```


## Plot

For plotting, you simply pipe the input data frame into heatmap, specifying:

- The rows, cols relative column names (mandatory)
- The value column name (mandatory)
- The annotations column name(s)

mtcars
```{r}
mtcars_heatmap <- 
	mtcars_tidy %>% 
		heatmap(`Car name`, Property, Value	) %>%
		add_tile(hp)

mtcars_heatmap
```

## Save

```{r eval=F}
mtcars_heatmap %>% save_pdf("mtcars_heatmap.pdf")
```

## Grouping

We can easily group the data (one group per dimension maximum, at the moment only the vertical dimension is supported) with dplyr, and the heatmap will be grouped accordingly

```{r}
mtcars_tidy %>% 
	group_by(vs) %>%
	heatmap(`Car name`, Property, Value	) %>%
	add_tile(hp)
```

## Custom palettes

We can easily use custom palette, using strings, hexadecimal color character vector, 

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = c("red", "white", "blue")
	)
```

Or a grid::colorRamp2 function for higher flexibility

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = circlize::colorRamp2(c(-2, -1, 0, 1, 2), viridis::magma(5))
	)
```

## Multiple groupings and annotations

```{r}
tidyHeatmap::pasilla %>%
	group_by(location, type) %>%
	heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`
		) %>%
	add_tile(condition) %>%
	add_tile(activation)
```

## Annotation types

**This feature requires >= 0.99.20 version**

"tile" (default), "point", "bar" and "line" are available

```{r}
# Create some more data points
pasilla_plus <- 
	tidyHeatmap::pasilla %>%
		dplyr::mutate(act = activation) %>% 
		tidyr::nest(data = -sample) %>%
		dplyr::mutate(size = rnorm(n(), 4,0.5)) %>%
		dplyr::mutate(age = runif(n(), 50, 200)) %>%
		tidyr::unnest(data) 

# Plot
pasilla_plus %>%
		heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`
		) %>%
	add_tile(condition) %>%
	add_point(activation) %>%
	add_tile(act) %>%
	add_bar(size) %>%
	add_line(age)
```

# Layer symbol

Add a layer on top of the heatmap

```{r}
tidyHeatmap::pasilla %>%
	
	# filter
	filter(symbol %in% head(unique(tidyHeatmap::pasilla$symbol), n = 10)) %>%
	
	heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`
		) %>% 
	layer_point(
		`count normalised adjusted log` > 6 & sample == "untreated3" 
	)
```
