---
title: "tidyHeatmap"
output: github_document
---

<!-- badges: start -->
  [![Lifecycle:maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->


```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)

library(magrittr)
library(dplyr)
library(tidyHeatmap)

```

Tidy heatmap. This package is a tidy wrapper of the package [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html). The goal of this package is to interface tidy data frames with this powerful tool.

Some of the advantages are:

- Row and/or columns colour annotations are easy to integrate just specifying one parameter (column names). 
- Custom grouping of rows is easy to specify providing a grouped tbl. For example `df %>% group_by(...)`
- Labels size adjusted by row and column total number
- Default use of Brewer and Viridis palettes



# Installation

To install the most up-to-date version

```{r, eval=FALSE}

devtools::install_github("stemangiola/tidyHeatmap")

```

To install the most stable version (however please keep in mind that this package is under a maturing lifecycle stage)

```{r, eval=FALSE}

install.packages("tidyHeatmap")

```

# Input data frame

Example of an input data frame

```{r}
tidyHeatmap::pasilla
```


# Plot

For plotting, you simply pipe the input data frame into heatmap, specifying:

- The horizontal, vertical relative column names (mandatory)
- The abundance column name (mandatory)
- The annotations column name(s)

```{r}
tidyHeatmap::pasilla %>%
	heatmap(
		.row = symbol,
		.column = sample,
		.value = `count normalised adjusted`,
		annotation = c(condition, type),
		transform = log1p
	)
```

# Grouping

We can easily group the data (one group per dimension maximum, at the moment only the vertical dimension is supported) with dplyr, and the heatmap will be grouped accordingly

```{r}
tidyHeatmap::pasilla %>%
	group_by(location, condition) %>%
	heatmap(
		.row = symbol,
		.column = sample,
		.value = `count normalised adjusted`,
		annotation = c(type),
		transform = log1p
	)
```

# Custom palettes

We can easily use custom palette, chooinga hexadecimal color character vector, or a grid::colorRamp2 functionfor higher flexibility

```{r}
pasilla %>%
	heatmap(
		.row = symbol,
		.column = sample,
		.value = `count normalised adjusted`,
		transform = log1p, 
		palette_abundance = circlize::colorRamp2(c(-2, -1, 0, 1, 2), viridis::magma(5))
	)
```