---
title: "Overview of the tidyHeatmap package"
author: "Stefano Mangiola"
date: "`r Sys.Date()`"
package: tidyHeatmap
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteEngine{knitr::knitr}
  %\VignetteIndexEntry{Overview of the tidyHeatmap package}
  %\usepackage[UTF-8]{inputenc}
---

[![Build Status](https://travis-ci.org/stemangiola/tidyHeatmap.svg?branch=master)](https://travis-ci.org/stemangiola/tidyHeatmap) [![Coverage Status](https://coveralls.io/repos/github/stemangiola/tidyHeatmap/badge.svg?branch=master)](https://coveralls.io/github/stemangiola/tidyHeatmap?branch=master)



```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, 
                      message = FALSE, cache.lazy = FALSE)

library(tidyverse)
library(tidyHeatmap)

```

This package is a tidy wrapper of the package [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html). The goal of this package is to interface a tidy data frame with this powerful tool.

Some of the advantages are:

- Row and/or columns colour annotations are easy to integrate just specifying one parameter (column names). 
- Custom grouping of rows is easy to specify providing a grouped tbl. For example `df %>% grup_by(...)`
- Labels size adjusted by row and column total number
- Default use of Brewer and Viridis palettes

## Input data frame

The input data frameha sto be a tidy `tbl` of this form

```{r}
pasilla
```

Where mandatory column are `sample`, `transcript`, count (named as you wish)


## Plot

For plotting, you simply pipe the input data frame into heatmap, specifying:

- The horizontal, vertical relative column names (mandatory)
- The abundance column name (mandatory)
- The annotations column name(s)

```{r}
pasilla %>%
	heatmap(
		.horizontal = sample,
		.vertical = symbol,
		.abundance = `count normalised adjusted`,
		annotation = c(condition, type),
		log_transform = T
	)
```

## Grouping

We can easily group the data (one group per dimension maximum, at the moment only the vertical dimension is supported) with dplyr, and the heatmap will be grouped accordingly

```{r}
pasilla %>%
	group_by(location) %>%
	heatmap(
		.horizontal = sample,
		.vertical = symbol,
		.abundance = `count normalised adjusted`,
		annotation = c(condition, type),
		log_transform = T
	)
```
# Appendix

```{r}
sessionInfo()
```
