---
title: "Overview of the tidyHeatmap package"
author: "Stefano Mangiola"
date: "`r Sys.Date()`"
package: tidyHeatmap
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of the tidyHeatmap package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, cache.lazy = FALSE)

library(magrittr)
library(dplyr)
library(tidyr)
library(tidyHeatmap)
library(purrr)

```

Tidy heatmap. This package is a tidy wrapper of the package [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html). The goal of this package is to interface tidy data frames with this powerful tool.

**Advantages:**

- Modular annotation with just specifying column names
- Custom grouping of rows is easy to specify providing a grouped tbl. For example `df %>% group_by(...)`
- Labels size adjusted by row and column total number
- Default use of Brewer and Viridis palettes

## Functions/utilities available

Function | Description
------------ | -------------
`heatmap` | Plot base heatmap
`add_tile` | Add tile annotation to the heatmap
`add_point` | Add point annotation to the heatmap
`add_bar` | Add bar annotation to the heatmap
`add_line` | Add line annotation to the heatmap
`save_pdf` | Save the PDF of the heatmap

## Installation

To install the most up-to-date version

```{r, eval=FALSE}

devtools::install_github("stemangiola/tidyHeatmap")

```

To install the most stable version (however please keep in mind that this package is under a maturing lifecycle stage)

```{r, eval=FALSE}

install.packages("tidyHeatmap")

```

## Input data frame

```{r}
mtcars_tidy = 
	mtcars %>% 
	as_tibble(rownames="Car name") %>% 
	
	# Scale
	mutate_at(vars(-`Car name`, -hp, -vs), scale) %>%
	
	# tidyfy
	gather(Property, Value, -`Car name`, -hp, -vs)

mtcars_tidy
```