---
title: "Overview of the tidyHeatmap package"
author: "Stefano Mangiola"
date: "`r Sys.Date()`"
package: tidyHeatmap
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Overview of the tidyHeatmap package}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, echo=FALSE, include=FALSE, }
library(knitr)
knitr::opts_chunk$set(cache = TRUE, warning = FALSE, message = FALSE, cache.lazy = FALSE)

library(magrittr)
library(dplyr)
library(tidyr)
library(tidyHeatmap)
library(purrr)

```

Tidy heatmap. This package is a tidy wrapper of the package [ComplexHeatmap](https://bioconductor.org/packages/release/bioc/html/ComplexHeatmap.html). The goal of this package is to interface tidy data frames with this powerful tool.

**Advantages:**

- Modular annotation with just specifying column names
- Custom grouping of rows is easy to specify providing a grouped tbl. For example `df %>% group_by(...)`
- Labels size adjusted by row and column total number
- Default use of Brewer and Viridis palettes

## Functions/utilities available

Function | Description
------------ | -------------
`heatmap` | Plot base heatmap
`add_tile` | Add tile annotation to the heatmap
`add_point` | Add point annotation to the heatmap
`add_bar` | Add bar annotation to the heatmap
`add_line` | Add line annotation to the heatmap
`save_pdf` | Save the PDF of the heatmap

## Installation

To install the most up-to-date version

```{r, eval=FALSE}

devtools::install_github("stemangiola/tidyHeatmap")

```

To install the most stable version (however please keep in mind that this package is under a maturing lifecycle stage)

```{r, eval=FALSE}

install.packages("tidyHeatmap")

```

## Contribution

If you want to contribute to the software, report issues or problems with the software or seek support pleae open an issue [here](https://github.com/stemangiola/tidyHeatmap/issues)

## Input data frame

The heatmaps visualise a multi-element, multi-feature dataset, annotated with independent variables. Each obsrvation is a element-feature pair (e.g., person-phisical characteristics).

element | feature | value | independent_variables
------------ | ------------- | ------------- | -------------
`chr` or `fctr` | `chr` or `fctr` | `numeric` | ...

Let's tranform the mtcars dataset into a tidy "element-feature-independent variables" data frame. Where the independent variables in this case are 'hp' and 'vs'.

```{r, echo=FALSE, include=FALSE}
library(dplyr)
library(tidyr)
library(tidyHeatmap)
```

```{r}
mtcars_tidy = 
	mtcars %>% 
	as_tibble(rownames="Car name") %>% 
	
	# Scale
	mutate_at(vars(-`Car name`, -hp, -vs), scale) %>%
	
	# tidyfy
	gather(Property, Value, -`Car name`, -hp, -vs)

mtcars_tidy
```


## Plot

For plotting, you simply pipe the input data frame into heatmap, specifying:

- The rows, cols relative column names (mandatory)
- The value column name (mandatory)
- The annotations column name(s)

mtcars
```{r}
mtcars_heatmap = 
	mtcars_tidy %>% 
		heatmap(`Car name`, Property, Value	) %>%
		add_tile(hp)

mtcars_heatmap
```

## Save

```{r eval=F}
mtcars_heatmap %>% save_pdf("mtcars_heatmap.pdf")
```

## Grouping

We can easily group the data (one group per dimension maximum, at the moment only the vertical dimension is supported) with dplyr, and the heatmap will be grouped accordingly

```{r}
mtcars_tidy %>% 
	group_by(vs) %>%
	heatmap(`Car name`, Property, Value	) %>%
	add_tile(hp)
```

## Custom palettes

We can easily use custom palette, using strings, hexadecimal color character vector, 

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = c("red", "white", "blue")
	)
```

Or a grid::colorRamp2 function for higher flexibility

```{r}
mtcars_tidy %>% 
	heatmap(
		`Car name`, 
		Property, 
		Value,
		palette_value = circlize::colorRamp2(c(-2, -1, 0, 1, 2), viridis::magma(5))
	)
```

## Multiple groupings and annotations

```{r}
tidyHeatmap::pasilla %>%
	group_by(location, type) %>%
	heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`
		) %>%
	add_tile(condition) %>%
	add_tile(activation)
```

## Annotation types

**This feature requires >= 0.99.20 version**

"tile" (default), "point", "bar" and "line" are available

```{r}
# Create some more data points
pasilla_plus = 
	tidyHeatmap::pasilla %>%
		dplyr::mutate(act = activation) %>% 
		tidyr::nest(data = -sample) %>%
		dplyr::mutate(size = rnorm(n(), 4,0.5)) %>%
		dplyr::mutate(age = runif(n(), 50, 200)) %>%
		tidyr::unnest(data) 

# Plot
pasilla_plus %>%
		heatmap(
			.column = sample,
			.row = symbol,
			.value = `count normalised adjusted`
		) %>%
	add_tile(condition) %>%
	add_point(activation) %>%
	add_tile(act) %>%
	add_bar(size) %>%
	add_line(age)
```

